Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-07-31 treestats_paper.tex
figure and more edits

Suppose we have sequenced a linear genome of $L$ bases from $n$ (homologous) sampled chromosomes,
and wish to compute a statistic of the allele frequency spectrum.
In other words, if the frequency of allele $a$ at a site $i$ is $p_i(a)$,
and the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $h(p)$,
then for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple average over sites, and monomorphic sites do not contribute,
we can compute this, summing over the $L$ loci, as
$$\sum_p h(p) f(p) = \frac{1}{L} \sum_{i=1}^L \sum_a f(p_i(a)).$$
(Note that since the ancestral allele is included, this is a function of the
\emph{folded} frequency spectrum, so is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and accounting for those that subsequently mutate).
What we do here is a generalization of this idea.

Roughly speaking, we will
(a) propagate ``weights'' additively up each tree,
(b) summarize these weights on each branch with a ``summary function'',
and (c) aggregate these summaries to produce statistics
(e.g., averaged in windows).

@@

A tree sequence describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient}.
These trees are highly correlated,
a fact which allows them to be stored and processed very efficiently,
as described in \citet{kelleher2018efficient}.
This is encoded as a sequence of trees, $\treeseq = (T_1, T_2, \ldots,
T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance the branch describing a relationship where $u$ is the child of $v$
is associated with $u$.
The length of this branch -- i.e., the difference in birth times of $v$ and $u$ --
we denote $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships
can cause the precise set of samples that lie below that edge to differ across
trees.

%%%%%%%%%%%%%%%%%%
\subsection*{Defining a statistic}

\jk{I think we're missing an overall overview of how to specify a statistic
and basic definitions of the framework itself. Before getting into the
specifics below, it would be good to give an more abstract idea of
what's happening. Perhaps with reference to a diagram?}

The basic ingredients that we use to summarize tree shape and genotype patterns
are a set of values associated with each sampled node, which we call the ``weights''.
For instance, we could calculate how many of a certain set $A$ of samples
are below a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree below that edge.
It turns out to be useful to allow more general weights (including negative ones).
These subtree-associated weights can be efficiently propagated along a tree sequence
because adjacent trees are simliar (and so common subtrees do not need to be updated).

We will use weights of the form described above to count numbers of samples in each subtree
frequently enough that we give them a special name --
for a subset set $A$ of samples,
the \emph{indicator weights of $A$} are the sample weights $\bone_A$ with
$\bone_A(u) = 1$ if $u \in A$ and $\bone_A(u) = 0$ otherwise.

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees below each node
% as we iterate through the tree sequence.

Next, we will describe how to use these subtree weights,
passed through a ``summary function'' in various ways,
to summarize genetic variation and tree shapes across the tree sequence.

\begin{definition}[Summary function]
    Given a weight vector $(w_1, \dots, w_k)$, the summary function
    $f: \mathbb{R}^k \rightarrow \mathbb{R}^m$ computes a vector
    $(\sigma_1, \dots, \sigma_m)$ summarising these weights.
\end{definition}

\jk{This definition is pretty weak, but wanted something in there.}

@@

Propagating indicator weights of a subset of samples up the tree
lets us easily count the number of those samples that inherit from each node in
the tree. Averaged across trees,
this tells us what proportion of the sample's genomes were inherited from each node.
Other choices of weights would tell us other information about the nodes.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree below node $u$,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)).
\end{align}
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

@@

@@

The first method of summarizing uses weights
to compute a summary for each site in the genome.
These are often reported by averaging the values over windows.
The general definition uses the \emph{allele weights} at each site in the genome,
defined simply to be the total weight of all nodes who have inherited that allele.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\section{misc stuff}

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Summary}
We have defined three types of statistic using the same weighting scheme:
\begin{enumerate}
    \item \textbf{Site statistics}
        -- summarizes weights below each allele at each polymorphic site.
    \item \textbf{Node statistics}
        -- summarizes total weight below each node.
    \item \textbf{Branch statistics}
        -- summarizes total weight below each branch, multiplied by branch length.
\end{enumerate}

\plr{TO-DO:}
\begin{itemize}

    \item What about normalization?
        In the initial implementation, Node statistics were divided by the length
        for which the node was an ancestor to any sample, rather than just total sequence length
        To make these more analogous to the other stats, we think we want to remove this.
        The denominator for the original case is easily computable, by using
        $f(x) = 1$ if $x>0$ and $f(x)=0$ otherwise, with appropriate (positive) weights.

\end{itemize}

Quick review of tree sequence terminology
and storage:
nodes,
edges,
sites,
mutations.

%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms for fast computation}

%%%%%%%%%%%%%%%%%%
\section*{Case studies}
% AKA "examples"

%%%%%%%%%%%%%%%%%%
\subsection*{Ancestry proportions}

%%%%%%%%%%%%%%%%%%
\subsection*{Population genetics statistics}

@@

{\Large \plr{OLD STUFF}}

%%%%%%%%%%%%%%%%%%
\subsection*{Efficient computation}

First we describe how to define
and efficiently compute a quite general set of functions of a collection of sequences
using tools from tree sequences.
In the next section we will specialize these to a natural set of statistics.

\plr{Since the ``weighting function' can take negtive values, maybe ``weight'' is not a good term.  Instead?}

\paragraph{Weighted site averages}
Suppose that we have $k$ groups of samples, $A_1$, \ldots, $A_k$.
% with $n_1$, \ldots, $n_k$ individuals in each, respectively.
The \emph{weight} we compute for a given site is a sum of weights computed for each allele at that site,
and the weights are found as a function of the vector of numbers of individuals in each group inheriting that allele.
Concretely, we specify a \emph{weighting function} $w(x_1, \ldots, x_k)$,
and we say that the \emph{average site weight} of a set of $L$ sites is
\begin{align} \label{eqn:average_site_weight}
    \site_w := \frac{1}{L} \sum_{i=1}^L \sum_{s \in I_i} w(x_1(s), \ldots, x_k(s)) ,
\end{align}
where $I_i$ is the set of alleles seen at that site and
$x_j(s)$ is the number of individuals in $A_j$ that carry allele $s$.

Before we describe how to compute this,
we give the analogous topological quantity,
that is equal to the expected weighted site average
under a continuum-sites model.

\paragraph{Weighted branch length averages}
The \emph{average branch length weight}
is computed almost just as above:
\begin{align} \label{eqn:average_branch_weight}
    \branch_w :=  \frac{1}{L} \sum_{i=1}^L \sum_{e \in T_i} \ell(e) w(x_1(e), \ldots, x_k(e)) ,
\end{align}
except that now $T_i$ is the genealogical tree at site $i$,
the sum is over edges $e$ in the tree, $\ell(e)$ is the length of the edge $e$,
and $x_j(e)$ is the number of individuals in $A_j$ that fall below $e$.

\emph{Note:} somewhat more generally,
we could assign a set of weights to the individuals in each group,
and let $x_j$ denote the total weight of the individuals in $A_j$
inheriting from that allele or edge.
For simplicity we do not pursue this here.
\plr{this is a good use of the word ``weight'' (although these could also be negative)}

\paragraph{Computation}
The \emph{tree differences} encoded by a tree sequence allow us to efficiently update the quantities $x$ used above.
As it is more straightforward, we first describe the algorithm
to compute the branch length version.

\plr{Rework these into three algorithms: one to maintain the sample weights $\iw$
    (basically already described as `tracked nodes' in msprime)
    and the other two that use this.}

\paragraph{Total branch length weight algorithm}
In a tree sequence, every branch in any tree is associated with its terminal \emph{node}.
We will keep updated an array $x$ so that for any node $e$,
the value $x[e][j]$ is the number of samples in $A_j$ below node $e$ in the current tree.
This will be zero for any node \emph{not} appearing in the current tree.
We also maintain $T$, the total weight of the current tree.
\begin{itemize}
    \item \emph{Initialize} all elements of $x$ to zero
        except that $x[a][j] = 1$ for each $a \in A_j$ and $1 \le j \le k$.
        Also set $S=0$ and $T=0$.
    \item Pop the next \emph{tree difference} off the stack: an interval with endpoints $\ell < r$,
        a set of edges $\mathcal{R}$ to remove, and a set of edges $\mathcal{A}$ to add.
    \item \emph{Remove} each edge (parent, child) in $\mathcal{R}$ from the tree,
        let $u$ = parent, and while $u != -1$,
        subtract $x[child]$ from $x[u]$ and set $u$ to be parent($u$).
    \item \emph{Add} each edge (parent, child) in $\mathcal{R}$ to the tree,
        let $u$ = parent, and while $u != -1$,
        add $x[child]$ to $x[u]$ and set $u$ to be parent($u$).
    \item For each node that has changed, let $w_\text{old}$ and $w_\text{new}$
        be the weights before and after the change in $x$,
        and add $w_\text{new} - w_text{old}$ to $T$.
    \item Add $T \times (r-\ell)$ to the running total, $S$.
    \item Return to (2).
\end{itemize}
\plr{should modify code to maintain a vector of $w$ values}
The average weighted branch length is obtained by dividing by the total number of sites.

The algorithm for sites is similar, also maintaining the array $x$,
and making use of the association of each mutation with a particular node in the tree.

\paragraph{Total site weight algorithm}
This works as the \emph{total branch length weight algorithm} except that steps (5--6) are replaced with:
\plr{gah this is terrible}
\begin{itemize}
    \item[5'] \emph{Find allele weights} for each mutated site in the tree:
        if $a$ is the ancestral allele,
        set $u[a]$ to be the value of $x$ at the root; and then,
        for each mutation at node $e$ with derived allele $b$
        and parent mutation occurring at node $p$ with derived allele $c$,
        add $x[e]$ to $u[b]$, and subtract $x[p]$ from $u[c]$.
    \item[6'] For each unique allele $a$ seen,
        add $w(v[a])$ to $T$.
\end{itemize}
\plr{check that algorithm deals correctly with more than one mut per node}

%%%%%%%%%%%%%%%%%%
\subsection*{General properties}

Here we describe some properties that we would like our statistics to have.

\paragraph{They do not depend on choice of reference allele.}
This is achieved by applying the \emph{same} weighting function to every allele.
Ancestral information can be incorporated by including an outgroup or ancestral sequence
as an additional group of size one.
Equivalently, we always work with \emph{unrooted} trees.

\paragraph{They do not depend on sample size.}
We'd like our statistics to be \emph{consistent estimators}
of a single quantity, and should therefore not depend on sample size in expectation.
\plr{insert this above}
This is achieved by always computing probabilies of sampling \emph{without replacement},
i.e., replacing each occurence of $p^k$ with $(x)_k/(n)_k$.
This lets us compare values of a given statistic
that are computed on different data sets with different sample sizes.

\paragraph{Equivalence of ``site'' and ``branch length'' statistics}
The definitions above are made so that for any weighting function,
$\site_w$ is a moment estimator of $\branch_w$, i.e.,
that $\E[\site_w] = \branch_w$ for \plr{somehow say continuum of sites here}.

>>>>

Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
-- i.e., the number of alleles whose frequency is $p$ in our sample --
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $h(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this is a function of the
is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and accounting for those that subsequently mutate).

What we do here is a generalization of this idea.
Roughly speaking, to do this we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics
(e.g., averaged in windows).
For instance, we could calculate how many of a certain set $A$ of samples
are below a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree below that edge.
It turns out to be useful to allow more general weights (including negative ones).
Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

\begin{figure}
    \centering
    \includegraphics{figures/divergence_diagram}
    \caption{
        An example of computing sequence divergence between red nodes and blue nodes
        from seven mutations at distinct sites in a sequence of length 10 using the tree.
        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
        a randomly chosen red node but not in a randomly chosen blue node.
        \textbf{(a)} Weights are propagated up the tree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, weights $(a,b)$ record how many red nodes ($a$)
        and how many blue nodes ($b$) are below each node.
        The weights at each node is the sum of the weights of all descendant sample nodes.
        \textbf{(c)} The summary function shown is applied to both the weight below 
        and the weight not below each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        The final value for divergence is the sum of these values divided by 10,
        the total number of sites.
        \label{fig:divergence_diagram}
    }
\end{figure}

We now begin to formalize these notions with definitions.

\begin{definition}[Summary function]
    For a set of $k$-dimensional weights with total weight $\tiw$,
    a summary function is a real-valued function $f(w_1, \ldots, w_k)$
    with the property that $f(0) = f(\tiw) = 0$.
\end{definition}

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.
The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
so that statistics do not depend on portions of the tree not ancestral to any of the samples.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees below each node
% as we iterate through the tree sequence.
In practice we allow summary functions to be vector-valued (so as to compute many statistics at once)
but this will not feature in our examples.
Next, we will describe how to use these subtree weights,
passed through a ``summary function'' in various ways,
to summarize genetic variation and tree shapes across the tree sequence.

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance the branch describing a relationship where $u$ is the child of $v$
is associated with $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships
can cause the precise set of samples that lie below that edge to differ across
trees.

@@

Perhaps the most natural way to summarise weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to 
count the number of samples that inherit from each node in the tree. 
Averaged across trees,
this tells us what proportion of the sample's genomes were inherited from each node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree below node $u$,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)).
\end{align}
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

We will use weights that count number of samples below each node
sufficiently often that we give them a special name --
for a subset set $A$ of samples,
the \emph{indicator weights of $A$} are the sample weights $\bone_A$ with
$\bone_A(u) = 1$ if $u \in A$ and $\bone_A(u) = 0$ otherwise.

@@

Node statistics only make sense in the context of the tree sequence,
but provide a useful bridge to the next type of statistic,
which are defined directly in terms of the genotypes
(and so do not depend on the tree sequence).

@@

The next mode of summarizing uses weights
to compute a summary for each site in the genome.
These are often reported by averaging the values over windows.
The general definition uses the \emph{allele weights} at each site in the genome,
defined simply to be the total weight of all nodes who have inherited that allele.

@@

@@
